---
title: "Course Project 1 - Reproducible Research"
author: "Roisin Kearins"
date: "17 September 2021"
output: html_document
---

```{r message=FALSE, results="hide"}
knitr::opts_chunk$set(echo = TRUE)
```
###Install and load packages

```{r message=FALSE, results="hide"}
install.packages("rmarkdown")
library(knitr)

#knitr::opts_chunk$set(error=true, echo = TRUE)
# Loading packages
library(ggplot2)
library(ggthemes)
library(dplyr)

```

### Read in dataset and then process it to format it correctly


```{r, echo=TRUE}
activity <- read.csv("activity.csv")

# Setting date format to help get the weekdays of the dates
activity$date <- as.Date(activity$date)
activity$interval <- as.factor(activity$interval)

# Getting the days of all the dates on the dataset

day <- weekdays(activity$date)

# Combining the dataset with the weekday of the dates
activity <- cbind(activity, day)

# Viewing the processed data
summary(activity)
```

####QUESTION 1
### Calculating total steps taken on a day
```{r, echo=TRUE}
activityTotalSteps <- with(activity, aggregate(steps, by = list(date), sum, na.rm = TRUE))


#### Changing col names
names(activityTotalSteps) <- c("Date", "Steps")

### Converting the data set into a data frame to be able to use ggplot2
totalStepsdf <- data.frame(activityTotalSteps)

####Plotting a histogram using ggplot2
png("graph1.png")
g <- ggplot(activityTotalSteps, aes(x = Steps)) + 
  geom_histogram(breaks = seq(0, 25000, by = 2500), fill = "#83CAFF", col = "black") + 
  ylim(0, 30) + 
  xlab("Total Steps Taken Per Day") + 
  ylab("Frequency") + 
  ggtitle("Total Number of Steps Taken on a Day")
dev.off()
print(g)
```
```{r}
mean(activityTotalSteps$Steps)
median(activityTotalSteps$Steps)
```

### QUESTION 2
### Calculating the average number of steps taken, averaged across all days by 5-min intervals.
```{r, echo=TRUE}
averageDailyActivity <- aggregate(activity$steps, by = list(activity$interval), FUN = mean, na.rm = TRUE)
### Changing col names
names(averageDailyActivity) <- c("Interval", "Mean")

### Converting the data set into a dataframe
averageActivitydf <- data.frame(averageDailyActivity)

### Plotting on ggplot2
png("graph2.png")

da <- ggplot(averageActivitydf, mapping = aes(Interval, Mean, group = 1)) + 
  geom_line(col = "purple") +
  xlab("Interval") + 
  ylab("Average Number of Steps") + 
  ggtitle("Average Number of Steps Per Interval") +
  theme_calc(base_family = "serif")

dev.off()
print(da)
```

```{r,echo=TRUE}

averageDailyActivity[which.max(averageDailyActivity$Mean), ]$Interval
```


### QUESTION 3
### Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs.

```{r,echo=TRUE}
sum(is.na(activity$steps))
### Matching the mean of daily activity with the missing values
imputedSteps <- averageDailyActivity$Mean[match(activity$interval, averageDailyActivity$Interval)]

### Transforming steps in activity if they were missing values with the filled values from above.
activityImputed <- transform(activity, 
                             steps = ifelse(is.na(activity$steps), yes = imputedSteps, no = activity$steps))

### Forming the new dataset with the imputed missing values.
totalActivityImputed <- aggregate(steps ~ date, activityImputed, sum)

### Changing col names
names(totalActivityImputed) <- c("date", "dailySteps")

sum(is.na(totalActivityImputed$dailySteps))

### Converting the data set into a data frame to be able to use ggplot2
totalImputedStepsdf <- data.frame(totalActivityImputed)

### Plotting a histogram using ggplot2
png("graph3.png")

p <- ggplot(totalImputedStepsdf, aes(x = dailySteps)) + 
  geom_histogram(breaks = seq(0, 25000, by = 2500), fill = "#83CAFF", col = "magenta") + 
  ylim(0, 30) + 
  xlab("Total Steps Taken Per Day") + 
  ylab("Frequency") + 
  ggtitle("Total Number of Steps Taken on a Day") + 
  theme_calc(base_family = "serif")
  
dev.off()
print(p)
```

```{r}
mean(totalActivityImputed$dailySteps)
median(totalActivityImputed$dailySteps)
```
### QUESTION4

### Are there differences in activity patterns between weekdays and weekends?

```{r}
###Updating format of the dates
activity$date <- as.Date(strptime(activity$date, format="%Y-%m-%d"))

### Creating a function that distinguises weekdays from weekends
activity$dayType <- sapply(activity$date, function(x) {
  if(weekdays(x) == "Saturday" | weekdays(x) == "Sunday")
  {y <- "Weekend"}
  else {y <- "Weekday"}
  y
})

### Creating the data set that will be plotted
activityByDay <-  aggregate(steps ~ interval + dayType, activity, mean, na.rm = TRUE)

### Plotting using ggplot
png("graph4.png")

dayPlot <-  ggplot(activityByDay, aes(x = interval , y = steps, color = dayType, group =1)) + 
  geom_line() + ggtitle("Average Daily Steps by Day Type") + 
  xlab("Interval") + 
  ylab("Average Number of Steps") +
  facet_wrap(~dayType, ncol = 1, nrow=2) +
  scale_color_discrete(name = "Day Type") +
  theme_calc(base_family = "serif")
dev.off()
print(dayPlot) 
```